Undecidability and the structure of the Turing degrees

### 2018-08-23 paper.tex
fixing typo
Implementing Ronno's suggestions

% \cite{soare16_turin_comput}
% \cite{soare1999history}
% \cite{lerman16:_degrees_unsol}
% \todo{}
\subsection{Decision Problems}
What does it mean to \emph{solve a problem}? What is a \emph{problem}, anyways?
These questions may seem esoteric, they must be addressed in order to build a
well defined theory of the difficulty and computability of problems. In this paper, we discuss a
specific type of problem, the \emph{decision problem}. Decision problems are
essentially sets of natural numbers, and to solve the problem is to be able to
decide whether any given element is inside the set or not. Because set theory is
commonly used as the foundational system of mathematics, we can convert problems
expressed in terms of more complex mathematical objects into decision problems,
and thus by studying decision problems can build a system that is able to assess
the computability of all mathematical problems. For example, we can express
strings over the alphabet $\{0,1\}$ as sets of natural numbers, and thus can
study problems about languages of such strings with the same tools we use for
decision problems. Going further, we can express the source code of a computer
program as a string over the alphabet $\{0,1\}$, and so can consider questions
about the behavior of computer programs.

\section{Models of Computation}
Upon consideration of different types of problems, it quickly becomes clear that
some are much more difficult than others. For example, when considering words
over the alphabet $\{a,b,c\}$, it seems much easier to determine whether a word
ends in an $a$ than whether a word is a palindrome. It seems even harder still
to determine whether the word is of the form $a^nb^nc^n$. In order to formalize
these notions of difficulty, we need to build abstract models of computation,
and then test their ability to decide such questions. We observe that a machine
that can solve the first problem only needs to ``remember'' the same amount of
information no matter how long the input is: the last letter of the word. In
contrast, the amount of memory that the second and third problems require is
dependent on the size of the input, because the first half of the palindrome or
$n$, respectively, can be arbitrarily large. The difference between the second
and third problems is more subtle: the second problem can be solved moving only
one direction in memory, while the third problem requires the ability to ``look
back'' in memory.
\par
Based on these differences, we can begin to construct different models of
computation that have just enough power to solve each problem. The first problem
can be solved by a \emph{\ac{FSA}}, a collection of a finite
number of states, a transition function, and a set of accepting states. The
\ac{FSA} accepts the input if the sequential application of each element to the
input finishes in an accepting state, and rejects otherwise. \aclp{FSA} are often represented by charts as in Figure~\ref{fig:fsa}, a
representation of an \ac{FSA} that decides the first problem.
\begin{figure}[h]
  \caption{An \ac{FSA} that determines whether a string ends in
    `$a$'.}
  \label{fig:fsa}
  \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
  \node[state,initial] (q_0) {$q_0$};
  \node[state,accepting] (q_1) [right=of q_0]{$q_1$};
  \path[->]
  (q_0) edge [loop above] node {b,c} ()
  edge [bend left] node  {a} (q_1)
  (q_1) edge [bend left] node {b,c} (q_0)
  edge [loop above] node {a} ();
\end{tikzpicture}
\end{figure}
\par
The second problem can be solved by a machine called a \emph{\ac{PDA}}, which is essentially a finite state automaton with the addition of
a stack. Rather than transforming the current state and the input into a new
state, as with a finite state automata, the transition function of a \ac{PDA} also considers the stack, which it can pop and push from. Within the
collection of \acp{PDA}, there is another division between
\emph{\acp{DPDA}} and \emph{Nondeterministic \aclp{PDA}}. For \acp{DPDA}, the
transition function only outputs one move, rather than a set of
moves. Nondeterministic \acp{PDA} are able to recognize more languages than
\acp{DPDA}. Figure~\ref{fig:pda} is a graphical representation of a \ac{PDA}
that decides palindromes. In this representation, $\epsilon$ is an empty input,
and  is a special symbol that denotes the bottom of the stack. The moves are
represented by a double of input and stack operation.

@@

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  Let $M$ be a Turing Machine and $i$ be an input. If $M$ is run on $i$, will it
  eventually halt?
  \begin{proof}[Proof of Undecidability]
    Suppose for the sake of contradiction that the halting problem is
    decidable. Then, there exists some machine
    $\mathcal{O}$ that can decide the halting problem. Then, we can construct a
    Turing machine $H$ that simulates $\mathcal{O}$ on its input $(M,i)$. In the
    case where $\mathcal{O}(M,i)$ accepts, $H$ enters into an infinite
    loop, whereas in the case where $\mathcal{O}(M,i)$ does not accept, $H$
    halts. Consider $H(H,\epsilon)$. If $H$ halts, then
    $\mathcal{O}(H,\epsilon)$ accepts, and so $H$ enters an infinite loop, and
    so it does not halt. But if $H$ does not halt, then
    $\mathcal{O}(H,\epsilon)$ does not accept, so $H$ halts. This is a
    contradiction, so our premise that the halting problem is decidable must be false.
  \end{proof}
\end{prob}
% \begin{thm}[Rice's Theorem] \cite{kozen99_autom}
%   % Proof by reduction to the halting problem?
% \end{thm}
% \subsection{Incompleteness}
% \begin{thm}[G\"odel's Incompleteness Theorem] 
%   % Kleene 1943? Also shown in hopcroft 1979
%   \cite{kleene43_recur_predic_quant}
% \end{thm}
\subsection{Reducibility}
\label{subsec:reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
  a machine $D_A$ that decides $A$.
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable~\cite{sipser13:_introd_theor_comput,post44:_recur,kleene80_introd}.

@@

\begin{defn}
  An oracle machine for a degree $A$, denoted $\mathcal{O}_A$, is a Turing
  machine equipped with a function that is able to decide $A$.
\end{defn}

@@

One of the earliest questions about the structure of the Turing degrees was
Post's Problem, first posed in 1944~\cite{post44:_recur}. Post's Problem asks
whether there is a computably enumerable degree strictly between $0$ and $0'$. It was solved
independently by Friedberg and Mucnik in the late 1950s using a technique
called the ``priority method''~\cite{Friedberg236, muchnik1956unsolvability}. A solution that does not
involve a priority argument was provided by Ku\v{c}era in 1986~\cite{Kucera:1986:APS:22416.22462}.

@@

\begin{lem}
  \label{lem:deg_count}
  Every Turing degree is of cardinality $\aleph_0$.
  \begin{proof}
    We have that since every Turing machine can be expressed as a finite string over
    $\{0,1\}$, it can be expressed as a finite subset of $\mathbb{N}$. So, it follows
    that for any given oracle, there are at most $\aleph_0$ oracle machines, so
    at most $\aleph_0$ problems are decidable for any Turing degree.
    \par
    Now that an upper bound has been established, we show that the cardinality
    of each degree is exactly $\aleph_0$. Suppose, for the sake of contradiction,
    that a Turing degree contained only a finite number of decision
    problems. Then, the degree would be decidable, since every finite set is
    decidable. But the only decidable degree is $0$, and in the previous section we
    showed that the cardinality of $0$ is $\aleph_0$, so it follows that every
    degree has cardinality $\aleph_0$.
  \end{proof}
\end{lem}

>>>>

% \cite{soare16_turin_comput}
% \cite{soare1999history}
% \cite{lerman16:_degrees_unsol}
% \todo{}
\subsection{Decision Problems}
What does it mean to \emph{solve a problem}? What is a \emph{problem}?
These questions may seem esoteric, they must be addressed in order to build a
well defined theory of the difficulty and computability of problems. In this paper, we discuss a
specific type of problem, the \emph{decision problem}. Decision problems are
essentially sets of natural numbers, and to solve the problem is to be able to
decide whether any given element is inside the set or not. Because set theory is
commonly used as the foundational system of mathematics, we can convert problems
expressed in terms of more complex mathematical objects into decision problems,
and thus by studying decision problems can build a system that is able to assess
the computability of all mathematical problems. For example, we can express
strings over the alphabet $\{0,1\}$ as sets of natural numbers, and thus can
study problems about languages of such strings with the same tools we use for
decision problems. Going further, we can express the source code of a computer
program as a string over the alphabet $\{0,1\}$, and so can consider questions
about the behavior of computer programs.

\subsection{Overview}
This paper aims to develop the ideas presented in the introduction in a natural
manner, and at a level of rigor accessible to an interested undergraduate. We
begin by presenting a few simple models of computation and assessing their
ability to compute various sets. We then introduce the Turing Machine, and
mention alternative models of equivalent power. Next, we introduce the concepts
of undecidability and Turing reducibility, and present the halting problem as an
example of an undecidable problem. In the interest of developing a strong
intuition regarding the concept of undecidability, we examine a variety of
decidable and undecidable problems about \aclp{CFG}. Finally, we employ the
Turing reduction to introduce a structure on the set of decision problems, and
present some basic results about its properties.

\section{Models of Computation}
Upon consideration of different types of problems, it quickly becomes clear that
some are much more difficult than others. For example, when considering words
over the alphabet $\{a,b,c\}$, it seems much easier to determine whether a word
ends in an $a$ than whether a word is a palindrome. It seems even harder still
to determine whether the word is of the form $a^nb^nc^n$. In order to formalize
these notions of difficulty, we need to build abstract models of computation,
and then test their ability to decide such questions. We observe that a machine
that can solve the first problem only needs to ``remember'' the same amount of
information no matter how long the input is: the last letter of the word. In
contrast, the amount of memory that the second and third problems require is
dependent on the size of the input, because the first half of the palindrome or
$n$, respectively, can be arbitrarily large. The difference between the second
and third problems is more subtle: the second problem can be solved moving only
one direction in memory, while the third problem requires the ability to ``look
back'' in memory.
\par
Based on these differences, we can begin to construct different models of
computation that have just enough power to solve each problem. The first problem
can be solved by a \emph{\ac{FSA}}, a collection of a finite
number of states, a transition function, and a set of accepting states. The
\ac{FSA} accepts the input if the sequential application of each element to the
input finishes in an accepting state, and rejects otherwise. \aclp{FSA} are often represented by diagrams as in Figure~\ref{fig:fsa}, a
representation of an \ac{FSA} that decides the first problem. An explanation of
the conventions used in these diagrams is available in Sipser's
\emph{Introduction to the Theory of Computation}~\cite[][pp. 34,114]{sipser13:_introd_theor_comput}
\begin{figure}[h]
  \caption{An \ac{FSA} that determines whether a string ends in
    `$a$'.}
  \label{fig:fsa}
  \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
  \node[state,initial] (q_0) {$q_0$};
  \node[state,accepting] (q_1) [right=of q_0]{$q_1$};
  \path[->]
  (q_0) edge [loop above] node {b,c} ()
  edge [bend left] node  {a} (q_1)
  (q_1) edge [bend left] node {b,c} (q_0)
  edge [loop above] node {a} ();
\end{tikzpicture}
\end{figure}
\par
The second problem can be solved by a machine called a \emph{\ac{PDA}}, which is essentially a finite state automaton with the addition of
a stack. Rather than transforming the current state and the input into a new
state, as with a finite state automata, the transition function of a \ac{PDA} also considers the stack, which it can pop and push from. Within the
collection of \acp{PDA}, there is another division between
\emph{\acp{DPDA}} and \emph{Nondeterministic \aclp{PDA}}. For \acp{DPDA}, the
transition function only outputs one move, rather than a set of
moves. Nondeterministic \acp{PDA} are able to recognize more languages than
\acp{DPDA}. Figure~\ref{fig:pda} is a graphical representation of a \ac{PDA}
that decides palindromes. In this representation, $\epsilon$ is an empty input,
and \$ is a special symbol that denotes the bottom of the stack. The moves are
represented by a double of input and stack operation.

@@

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  Let $M$ be a Turing Machine and $i$ be an input. If $M$ is run on $i$, will it
  eventually halt?
  \begin{proof}[Proof of Undecidability]
    Suppose for the sake of contradiction that the halting problem is
    decidable. Then, there exists some machine
    $\mathcal{O}$ that can decide the halting problem. Then, we can construct a
    Turing machine $H$ that simulates $\mathcal{O}$ on its input $(M,i)$. In the
    case where $\mathcal{O}(M,i)$ accepts, $H$ enters into an infinite
    loop, whereas in the case where $\mathcal{O}(M,i)$ does not accept, $H$
    halts. Consider $H(H,\epsilon)$. If $H$ halts, then
    $\mathcal{O}(H,\epsilon)$ accepts, and so $H$ enters an infinite loop, and
    so it does not halt. But if $H$ does not halt, then
    $\mathcal{O}(H,\epsilon)$ does not accept, so $H$ halts. This is a
    contradiction, so our premise that the halting problem is decidable must be false.
  \end{proof}
\end{prob}
% \begin{thm}[Rice's Theorem] \cite{kozen99_autom}
%   % Proof by reduction to the halting problem?
% \end{thm}
% \subsection{Incompleteness}
% \begin{thm}[G\"odel's Incompleteness Theorem] 
%   % Kleene 1943? Also shown in hopcroft 1979
%   \cite{kleene43_recur_predic_quant}
% \end{thm}
\subsection{Reducibility}
\label{subsec:reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
  to $B$ and write $A \leq_T B$ if it is possible to construct an oracle
  machine for $B$ that decides $A$ (see
  Definition~\ref{defn:oracle}).
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable~\cite{sipser13:_introd_theor_comput,post44:_recur,kleene80_introd}.

@@

\begin{defn}
  \label{defn:oracle}
  An oracle machine for a degree $A$, denoted $\mathcal{O}_A$, is a Turing
  machine equipped with a function that is able to decide $A$.
\end{defn}

@@

One of the earliest questions about the structure of the Turing degrees was
Post's Problem, first posed in 1944~\cite{post44:_recur}. Post's Problem asks
whether there is a computably enumerable degree strictly between $0$ and $0'$. A
positive answer to the question was found independently by Friedberg and Mucnik in the late
1950s using a technique called the ``priority method''~\cite{Friedberg236,
  muchnik1956unsolvability}. A solution that does not involve a priority
argument was provided by Ku\v{c}era in 1986~\cite{Kucera:1986:APS:22416.22462}.

@@

\begin{lem}
  \label{lem:deg_count}
  Every Turing degree is of cardinality $\aleph_0$.
  \begin{proof}
    We have that since every Turing machine can be expressed as a finite string over
    $\{0,1\}$, it can be expressed as a finite subset of $\mathbb{N}$. So, it follows
    that for any given oracle, there are at most $\aleph_0$ oracle machines, so
    at most $\aleph_0$ problems are decidable for any Turing degree.
    \par
    Now that an upper bound has been established, we show that the cardinality
    of each degree is exactly $\aleph_0$. Suppose, for the sake of contradiction,
    that a Turing degree contained only a finite number of decision
    problems. Then, the degree would be decidable, since every finite set is
    decidable. But we know that the only decidable degree is $0$, since all
    decidable problems are mutually reducibile, and in the previous section we
    showed that the cardinality of $0$ is $\aleph_0$, so it follows that every
    degree has cardinality $\aleph_0$.
  \end{proof}
\end{lem}
